---
title: "Data 511, Analyzing NFL Seasons"
author: "Kristofer Hormoz"
date: ""
output:
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Theses are all the packages used in this project.

library(tinytex)

library(caret)

library(psych)

library(ggplot2)

library(gridExtra)

library(rattle)

library(plyr)

library(RANN)

library(rpart)

library(rpart.plot)
```{r error=TRUE, include = FALSE}
# error = TRUE means to ignore the error messages
# packages used --------------------------------------------------------------
library(tinytex)
library(caret)
library(psych)
library(ggplot2)
library(gridExtra)
library(rattle)
library(plyr)
library(RANN)
library(rpart)
library(rpart.plot)
```


In this project I will be analyzing some stats from the 2020 NFL season. 
In doing so we will go over everything that we did in Data 511.

I will be getting my data from https://www.pro-football-reference.com/

**Section 1: Data Preparation Phase**

This is the 2020 nfl season data set
```{r}
nfl_2020_season <- read.csv("nfl_2020_season.csv")
colnames(nfl_2020_season) <- nfl_2020_season[1,] # had to fix the column names
nfl_2020_season <- nfl_2020_season[-c(1),] # deleted a row 

summary(nfl_2020_season[,c(1:5)]) # just looking at first 5 variables
# all of the variables are char when they should be numeric, we need to fix this
```




We will now fix some of the variables. Note that we will keep "Tm" which
stands for team as a char variable. Variables 16 to 28 will be removed.
```{r, include=FALSE}
nfl_2020_season$Rk <- as.numeric(as.character(nfl_2020_season$Rk)) #converting 
                                                              # rank to numeric

nfl_2020_season$PF <- as.numeric(as.character(nfl_2020_season$PF))

nfl_2020_season[,c(4)]<- as.numeric(as.character(nfl_2020_season[,c(4)]))

nfl_2020_season[,c(5)]<- as.numeric(as.character(nfl_2020_season[,c(5)]))

# or use a for loop

for (i in c(3:21)){
  nfl_2020_season[,c(i)]<-
  as.numeric(as.character(nfl_2020_season[,c(i)]))
}

summary(nfl_2020_season)
```




These will be the variables we will be analyzing in our project.

The variables used are;

1) RK: symbolizes team rank

2) Tm: symbolizes the team

3) PF: symbolizes points fought for ( i.e total points gained in a season)

4) Yds: symbolizes yards gained by passing

5) Ply: symbolizes offensive plays (pass attempts + Rush attempts + Times Sacked)

6) Y/p: symbolizes yards per play

7) TO: symbolizes Turn overs lost

8) FL: Symbolizes fumbles lost

9) 1std: Symbolizes first downs gained

10) cmp: Symbolizes completions

11) Att: symbolizes attempts made by passing 

12) Yds.1: Symbolizes Total yards gained by rushing

13) TD: symbolizes touchdowns

14) Int: Symbolizes interceptions thrown

15) Yds.1: Symbolizes total yards from passing

16) Yds.2: Symbolizes total yards made by rushing

24) year: symbolizes year
```{r}
# now we will fix up pur data set and only keep variables we are interested in

nfl_2020_season <- nfl_2020_season[,c(1:21)]

nfl_2020_season$year <- 2020 # adding year 2020
nfl_2020_season$G <- NULL
str(nfl_2020_season) # str gives the structure of each variable.
```




The next lines of code, we are downloading more data and preparing it. 
Seasons 2016 to 2019 will be added.
```{r}
nfl_2019_season <- read.csv("nfl_2019_season.csv")

colnames(nfl_2019_season) <- nfl_2019_season[1,] # had to fix the column names
nfl_2019_season <- nfl_2019_season[-c(1),] # deleted a row 

nfl_2018_season <- read.csv("nfl_2018_season.csv")

colnames(nfl_2018_season) <- nfl_2018_season[1,] # had to fix the column names
nfl_2018_season <- nfl_2018_season[-c(1),] # deleted a row 

nfl_2017_season <- read.csv("nfl_2017_season.csv")

colnames(nfl_2017_season) <- nfl_2017_season[1,] # had to fix the column names
nfl_2017_season <- nfl_2017_season[-c(1),] # deleted a row 

nfl_2016_season <- read.csv("nfl_2016_season.csv")

colnames(nfl_2016_season) <- nfl_2016_season[1,] # had to fix the column names
nfl_2016_season <- nfl_2016_season[-c(1),] # deleted a row =
```




```{r, include=FALSE}
# changing the variables for the 2019 season.

nfl_2019_season$Rk <- as.numeric(as.character(nfl_2019_season$Rk)) 

nfl_2019_season$PF <- as.numeric(as.character(nfl_2019_season$PF))

for (i in c(3:21)){
  nfl_2019_season[,c(i)]<-
  as.numeric(as.character(nfl_2019_season[,c(i)]))
}

nfl_2019_season <- nfl_2019_season[c(1:32),c(1:21)]

nfl_2019_season$year <- 2019 # adding year 2019 
nfl_2019_season$G <- NULL
summary(nfl_2019_season)
```



```{r,include=FALSE}

# changing the variables for the 2018 season.

nfl_2018_season$Rk <- as.numeric(as.character(nfl_2018_season$Rk)) 

nfl_2018_season$PF <- as.numeric(as.character(nfl_2018_season$PF))

for (i in c(3:21)){
  nfl_2018_season[,c(i)]<-
  as.numeric(as.character(nfl_2018_season[,c(i)]))
}

nfl_2018_season <- nfl_2018_season[c(1:32),c(1:21)]

nfl_2018_season$year <- 2018
nfl_2018_season$G <- NULL
summary(nfl_2018_season)
```


```{r,include=FALSE}
# changing the variables for the 2017 season.

nfl_2017_season$Rk <- as.numeric(as.character(nfl_2017_season$Rk)) 

nfl_2017_season$PF <- as.numeric(as.character(nfl_2017_season$PF))

for (i in c(3:21)){
  nfl_2017_season[,c(i)]<-
  as.numeric(as.character(nfl_2017_season[,c(i)]))
}

nfl_2017_season <- nfl_2017_season[c(1:32),c(1:21)]

nfl_2017_season$year <- 2017
nfl_2017_season$G <- NULL
summary(nfl_2017_season)
```


```{r,include=FALSE}
# changing the variables for the 2016 season.

nfl_2016_season$Rk <- as.numeric(as.character(nfl_2016_season$Rk)) 

nfl_2016_season$PF <- as.numeric(as.character(nfl_2016_season$PF))

for (i in c(3:21)){
  nfl_2016_season[,c(i)]<-
  as.numeric(as.character(nfl_2016_season[,c(i)]))
}

nfl_2016_season <- nfl_2016_season[c(1:32),c(1:21)]

nfl_2016_season$year <- 2016
nfl_2016_season$G <- NULL
summary(nfl_2016_season)
```

**Combining the Data into One Data Frame**

Here we are merging the data sets by the rows into one data frame. This will
Give a new data frame "nfl" with 160 observations with 15 variables.
```{r}
nfl <- rbind(nfl_2020_season[,c(21,1:20)],nfl_2019_season[,c(21,1:20)],
             nfl_2018_season[,c(21,1:20)],nfl_2017_season[,c(21,1:20)],
             nfl_2016_season[,c(21,1:20)])

nfl$pass_Att <- nfl$Att
nfl$rush_Att <- nfl$Att.1
nfl$pass_TD <- nfl$TD
nfl$rush_TD <- nfl$TD.1

nfl$Att <- NULL
nfl$Att.1 <- NULL
nfl$TD <- NULL
nfl$TD.1 <- NULL

```


**Exploratory Data Analysis Phase**



*step (2) using decision trees to find optimal bins*

```{r}
# Normalized histogram w churn overlay.

grid.arrange(
ggplot(nfl, aes(pass_Att)) + 
  geom_histogram(aes(fill = pass_TD), 
                 color = "black", binwidth= 20, position = "fill") + 
  xlab("Passing Attempts Made") + 
  ylab("Touchdowns") + 
  ggtitle("TD Made By Passing"),



# Non-normalized histogram w overlay.
ggplot(nfl, aes(pass_Att)) + 
  geom_histogram(aes(fill = pass_TD), 
                 color = "black", binwidth = 1, position = "stack") + 
  xlab("Passing Attempts Made") + 
  ylab("Touchdowns") + 
  ggtitle("TD Made by Passing"),


ggplot(nfl, aes(rush_Att)) + 
  geom_histogram(aes(fill = rush_TD), 
                 color = "black", binwidth= 20, position = "fill") + 
  xlab("Rushing Attempts Made") + 
  ylab("Touchdowns") + 
  ggtitle("TD Made By Rushing"),



# Non-normalized histogram w overlay.
ggplot(nfl, aes(rush_Att)) + 
  geom_histogram(aes(fill = rush_TD), 
                 color = "black", binwidth = 1, position = "stack") + 
  xlab("Rushing Attempts Made") + 
  ylab("Touchdowns") + 
  ggtitle("TD Made by Rushing")

)

nfl$pass_TD_cut <- cut(nfl$pass_TD, breaks = c(0,30,40,100))
nfl$rush_TD_cut <- cut(nfl$rush_TD, breaks = c(0,10,20,30))

nfl$pass_Att_cut <- cut(nfl$pass_Att, breaks = c(0,500,600,700))
nfl$rush_Att_cut <- cut(nfl$rush_Att, breaks = c(0,400,475,600))

nfl$pass_Att_cut <- as.factor(nfl$pass_Att_cut)
nfl$rush_Att_cut <- as.factor(nfl$pass_Att_cut)
nfl$pass_TD_cut <- as.factor(nfl$pass_TD_cut)
nfl$rush_TD_cut <- as.factor(nfl$rush_TD_cut)

grid.arrange(
# non normalized
ggplot(nfl, aes(pass_Att_cut))+ 
  geom_bar(aes(fill = pass_TD_cut),position = "stack")+ 
  xlab("Passing Attempts Made")+
  ylab("Touchdowns Made"),

#normalized
ggplot(nfl, aes(pass_Att_cut))+ 
  geom_bar(aes(fill = pass_TD_cut) ,position = "fill")+ 
  xlab("Passing Attempts Made")+
  ylab("Touchdowns Made"),

ggplot(nfl, aes(rush_Att_cut))+ 
  geom_bar(aes(fill = rush_TD_cut),position = "stack")+ 
  xlab("Rushing Attempts Made")+
  ylab("Touchdowns Made"),

#normalized
ggplot(nfl, aes(rush_Att_cut))+ 
  geom_bar(aes(fill = rush_TD_cut) ,position = "fill")+ 
  xlab("Rushing Attempts Made")+
  ylab("Touchdowns Made")

)

grid.arrange(

  ggplot(data = nfl,
       aes(x = pass_Att_cut, y = pass_TD, color = pass_Att_cut)) +
  geom_boxplot(size = 1.2, alpha = .5) +
  xlab("Passing Attempts Made") +
  ylab("Touchdowns Made By Passing") +
  labs(title = "Boxplot of TD's Made By Passing Attempts") +
  scale_color_brewer(palette="Dark2") +
  geom_jitter(shape=16, position=position_jitter(0.2)),


ggplot(data = nfl,
       aes(x = rush_Att_cut, y = rush_TD, color = rush_Att_cut)) +
  geom_boxplot(size = 1.2, alpha = .5) +
  xlab("Rushing Attempts Made") +
  ylab("Touchdowns Made By Rushing") +
  labs(title = "Boxplot of TD's Made By Rushing Attempts") +
  scale_color_brewer(palette="Dark2") +
  geom_jitter(shape=16, position=position_jitter(0.2))

)
# then do a contingency table

t1 <- table(nfl$pass_TD_cut, nfl$pass_Att_cut)
round(prop.table(t1,2)*100,2)


t2 <- table(nfl$rush_TD_cut, nfl$rush_Att_cut)
round(prop.table(t2,2)*100,2)

```



